The illusory promise of the Aligned Rank Transform

### 2025-12-29 sandbox.qmd
still updating the appendix

```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
library(dplyr)
library(tidyverse)
source("utils-data-reading.R")
source("utils-plotlying.R")
```

### Statistical power {#power}
We evaluate the statistical power of the four methods on the three two-factor designs. Because there is a tradeoff between Type I and Type II errors, high power may simply reflect an inflated Type I error rate. Since a main effect on $X_1$ inflates ART's Type I error rates for $X_2$ and for their interaction (see previous sections), we therefore focus on single-effect scenarios.

**Main effects**. We vary the magnitude of effect on $X_1$ and $X_2$ separately and assess each method's power to detect the corresponding effect at $\alpha = .05$. Specifically, we manipulate the parameters $a_1$ or $a_2$ over the range $0.2$ to $0.8$. Depending on the experimental design, this range allows us to simulate both low-power settings (e.g., power below $30\%$) and high-power settings (e.g., power above $80\%$).

@fig-power-main-1 presents the results for $X_1$. In many configurations, differences in power across methods are modest. To facilitate comparison, we also report method rankings; readers are encouraged to interact with the figures for precise power estimates.

PAR achieves the highest power under the normal and exponential distributions, but its power is particularly low under the log-normal distribution. ART initially appears advantageous under the the log-normal, binomial, and ordinal distributions for small effects, likely reflecting inflated Type I errors under these conditions. As $a_1$ grows larger, ART's relative advantage diminishes, and it ultimately becomes the least powerful method. Finally, INT consistently outperforms RNK and ranks highly across all conditions.

::: {#fig-power-main-1}
<div id="plot-power-1.1">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Power"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

data <- read_data(prefix, alpha = .05, effectType = 3, distributions)
df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX1", max = 104,  ytitle = 'Power (%)')
```
</div>

<div id="plot-power-1.2">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power - ranking')
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-1.2").classList.add("hidden"); }); </script>

<button id="btn1" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-1.1');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-1.2').classList.toggle('hidden');
  btn1.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **main effect of $X_1$** as a function of its magnitude $a_1$, with $n=20$.
:::

@fig-power-main-2 presents our results for $X_2$. Because this factor has a different number of levels across designs, we assess different ranges of power for each design. Nevertheless, the same overall patterns emerge.

::: {#fig-power-main-2}
<div id="plot-power-2.1">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
data <- read_data(prefix, alpha=.05, effectType = 4, distributions) # CHANGE n TO 20

df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX2", max = 104,  ytitle = 'Power (%)')
```
</div>

<div id="plot-power-2.2">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power - ranking')
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-2.2").classList.add("hidden"); }); </script>

<button id="btn2" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-2.1');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-2.2').classList.toggle('hidden');
  btn2.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **main effect of $X_2$** as a function of its magnitude $a_2$, with $n=20$. 
:::

**Interaction effects**. We also vary the magnitude of the interaction effect $X_1 \times X_2$ by manipulating the parameter $a_{12}$ over the range $0.5$ to $2.0$, and then measure the power to detect this interaction. The results, summarized in @fig-power-interaction, are consistent with the patterns observed for main effects.

::: {#fig-power-interaction}
<div id="plot-power-3.1">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
data <- read_data(prefix, alpha=.05, effectType = 5, distributions)

df <- reshape_by_design(data, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_2(df, xlab = "magnitude of interaction effect", var = "rateX1X2", xvar = "effectX1X2", max = 104,  ytitle = 'Power (%)')
```
</div>

<div id="plot-power-3.2">
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- data %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power - ranking')
```
</div> <style> .hidden { display: none; } </style> <script> document.addEventListener("DOMContentLoaded", function() { document.getElementById("plot-power-3.2").classList.add("hidden"); }); </script>

<button id="btn3" style="background-color: #f0f0f0; border: 1px solid #ccc; color: #333; font-size: 0.9em;
    padding: 4px 10px; border-radius: 6px; cursor: pointer; float: right;"
  onclick="const plot = document.getElementById('plot-power-3.1');
  plot.classList.toggle('hidden');
  document.getElementById('plot-power-3.2').classList.toggle('hidden');
  btn3.innerText = plot.classList.contains('hidden') ? 'show percent' : 'show ranking';
">
  show ranking
</button>

Power ($\alpha = .05$) for detecting the **interaction effect $X_1 \times X_2$** as a function of its magnitude $a_{12}$, with $n=20$.
:::

Overall, there is no clear winner, but we observe that PAR shows poor power performance for log-normal data, while INT consistently outperforms RNK. Furthermore, ART provides no substantial power advantage, despite its inflated Type I error rates.

>>>>

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(tidyverse)

source("utils-data-reading.R")
source("utils-plotlying.R")
```

We present complementary results and new experiments that investigate additional scenarios. We also compare INT and RNK with other nonparametric methods. Unless explicitly mentioned in each section, we follow the experimental methodology presented in the main article. At the end of each section, we summarize our conclusions.

## ANOVA-type statistic (ATS) {#ATS}
We compare PAR, RNK, and INT to the ANOVA-type statistic (ATS) [@Brunner_ATS:2001] for two-factor designs. We use its implementation in the R package *nparLD* [@nparLD], which does not support between-subjects designs. Thus, we only evaluate it for the $4 \times 3$ within-subjects and the $2 \times 4$ mixed designs.

**Type I error rates: Main effects**. @fig-ATS-designs-main presents Type I error rates for the main effect of $X_2$. Under the mixed design, RNK, INT, and ATS exhibit very similar error rates, which are close to nominal levels. In the within-subjects design, the error rates of ATS tend to be slightly above $5\%$. Additionally, unlike the other methods whose error rates drop significantly below $5\%$ when the effect of $X_1$ becomes stronger under binomial and ordinal scales, the power of ATS does not seem to be affected in these cases.

::: {#fig-ATS-designs-main}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
prefix <- "Type_I_ATS"
alpha = 0.05

distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")
methods = c("PAR", "RNK", "INT", "ATS")
palette = c("#888888", "#E69F00", "#009E73", "#FF70AB")

df <- read_data(prefix, alpha, effectType = 1, distributions, methods = methods)
df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotly_error_by_design_3(df, xlab = "magnitude of main effect", var = "rateX2", xvar = "effectX1", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_2$** as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$)
:::

@fig-ATS-designs-main-2 presents results for the main effect of $X_1$. The error rates of ATS are now slightly inflated under the mixed design. The other methods exhibit the same trends as for the other factor.

::: {#fig-ATS-designs-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df2 <- read_data(prefix, alpha, effectType = 2, distributions, methods = methods)
df2 <- reshape_by_design(df2, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

plotly_error_by_design_3(df2, xlab = "magnitude of main effect", var = "rateX1", xvar = "effectX2", max = 7.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **main effect of $X_1$** as a function of the magnitude $a_2$ of the main effect of $X_2$ ($n = 20$)
:::

**Type I error rates: Interactions**. @fig-ATS-designs-interaction-1 presents Type I error rates for the interaction in the presence of a single main effect. Results are again very similar for all three nonparametric methods under the mixed design. In contrast, the error rates of ATS tend to be lower than nominal levels under the within-subjects design, often falling below $4\%$.

::: {#fig-ATS-designs-interaction-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
plotly_error_by_design_3(df, xlab = "magnitude of main effect", var = "rateX1X2", xvar = "effectX1", max = 10.1, nticks=8, cbPalette = palette)
```
Type I error rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitude $a_1$ of the main effect of $X_1$ ($n = 20$) 
:::

 When two parallel main effects are present, ATS and RNK lead to very similar trends (see @fig-ATS-designs-interaction-2). Overall, INT appears to be a more robust method with the exception of the binomial distribution, for which error rates are higher for this method.

::: {#fig-ATS-designs-interaction-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- read_data(prefix, alpha, effectType = 0, distributions, methods = methods)
df <- reshape_by_design(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_error_by_design_3(df, xlab = "magnitude of main effects", ytitle = 'Positives (%)', var = "rateX1X2", xvar = "effectX1", max = 104, nticks=8, cbPalette = palette)
```
Positive rates ($\alpha = .05$) for the **interaction** $X_1 \times X_2$ as a function of the magnitudes $a_1 = a_2$ of the main effects of $X_1$ and $X_2$  ($n = 20$). *Note: These values correspond to Type I error rates under the null hypothesis $a_{12} = 0$; alternative definitions of the null hypothesis may lead to different results.* 
:::

**Power: Main effects**. As shown in @fig-ATS-power-main-1 and @fig-ATS-power-main-2, ATS appears as the most powerful method for detecting effects of $X_1$ for the mixed design. However, in all other situations, it has less power than INT and offer less power than RNK.

::: {#fig-ATS-power-main-1}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
alpha <- .05

prefix <- "Power_ATS"
distributions = c("norm", "lnorm", "exp", "poisson", "binom", "likert")
dnames = c("Normal", "Log-normal", "Exponential", "Poisson", "Binomial", "Ordinal (5 levels)")

df <- read_data(prefix, alpha, effectType = 3, distributions, methods = methods)

df <- df %>% arrange(design,family,effectX1,rateX1)  %>% group_by(design,family,effectX1) %>% mutate(rank = rank(rateX1))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design_3(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX1", xvar = "effectX1", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_1$** as a function of the magnitude of effect $a_1$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

::: {#fig-ATS-power-main-2}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- read_data(prefix, alpha, effectType = 4, distributions, methods = methods)

df <- df %>% arrange(design,family,effectX2,rateX2)  %>% group_by(design,family,effectX2) %>% mutate(rank = rank(rateX2))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design_3(df, xlab = "magnitude of main effect", var = "rank", hovervar = "rateX2", xvar = "effectX2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **main effect of $X_2$** as a function of the magnitude of effect $a_2$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

**Power: Interactions**. @fig-ATS-power-interaction shows results on power for interactions. INT emerges again as the most powerful method. The power of ATS is particularly low under the within-subjects design.

::: {#fig-ATS-power-interaction}
```{r, echo=FALSE, fig.height=3.3, fig.width = 9, warning=FALSE}
df <- read_data(prefix, alpha, effectType = 5, distributions, methods = methods)

df <- df %>% arrange(design,family,effectX1X2,rateX1X2)  %>% group_by(design,family,effectX1X2) %>% mutate(rank = rank(rateX1X2))
df <- as.data.frame(df) %>% reshape_by_design(dnames, effectvars = c("effectX1","effectX2","effectX1X2"))

#df <- reshapeByDesign(df, dnames, effectvars = c("effectX1","effectX2","effectX1X2"))
plotly_power_by_design_3(df, xlab = "magnitude of interaction effect", var = "rank", hovervar = "rateX1X2", xvar = "effectX1X2", max = 4.2, ytitle = 'Power (%) - ranking', cbPalette = palette)
```
Ranking of methods for their power ($\alpha = .05$) to detect the **interaction effect $X_1 \times X_2$** as a function of the magnitude of effect $a_{12}$ ($n=20$). Hover over the charts to inspect their power ($\%$).
:::

### Conclusion
Although ATS appears to be a valid alternative, it does not offer clear performance advantages over INT, which is also simpler and more versatile.

### 2025-12-29 text/30-assumptions.qmd
still updating the appendix

However, a question remains: Is ART’s failure in handling such models due solely to heteroscedasticity issues? For example, consider the two time distributions in @fig-homoscedastic-model, which --- despite having different shapes --- share the same standard deviation. Would ART perform correctly in this case?

::: {#fig-homoscedastic-model}
```{r, echo=FALSE, message=FALSE, fig.height=2.5, fig.width = 9, warning=FALSE}
library(plotly)

sdlog <- function(mean, sd) {
	sqrt(log(1 + sd^2/mean^2))	
}

meanlog <- function(mean, sd) {
	log(mean^2/sqrt(sd^2 + mean^2))
}

xs <- seq(0, 10, length.out = 1000)

sd <- 1 # This is the common standard deviation
m1 <- 0.8
m2 <- 1.4

logm1 <- meanlog(m1, sd)
logm2 <- meanlog(m2, sd)

logsd1 <- sdlog(m1, sd)
logsd2 <- sdlog(m2, sd)

y1s <- dlnorm(xs, logm1, logsd1)
y2s <- dlnorm(xs, logm2, logsd2)

x1 <- 0.36
x2 <- 1.30

annot1 <- paste("Log-normal: mean = ", m1, ", sd = ", sd)
annot2 <- paste("Log-normal: mean = ", m2, ", sd = ", sd)

fig <- plot_ly() %>%  
      add_lines(x = xs, y = y1s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot1, hoverinfo = 'text', line=list(color=palette[1])) %>%
      add_lines(x = xs, y = y2s, visible = TRUE, type = 'scatter', mode = 'lines', text = annot2, hoverinfo = 'text', line=list(color=palette[2])) %>%      
      add_annotations(x = x1, y = max(y1s), text = "easy", xref = "x", yref = "y", showarrow = F, font = list(color = palette[1]), xanchor = 'left') %>% 
      add_annotations(x = x2, y = max(y2s), text = "hard", xref = "x", yref = "y", showarrow = F, font = list(color = palette[2]), xanchor = 'left') %>%   
      layout(
        showlegend = FALSE,
        xaxis = list(title = "Time (min)", range = c(0, 4.5), showgrid = F, showticks = T, ticks="outside", zeroline = F), 
        yaxis = list(showgrid = F, showticklabels = F, showline = F, range=c(0, max(y1s) + .05), fixedrange=T)
      ) %>% 
	      config(displayModeBar = TRUE, scrollZoom = FALSE, displaylogo = FALSE, modeBarButtonsToRemove = c("lasso2d", "select2d",  "zoomIn2d", "zoomOut2d", "autoscale", "hoverclosest", 'hoverCompare'))  %>% layout(dragmode='pan')

```
Log-normal distributions with equal variances but different means. Notice that their shapes are distinct. 
:::

While we do not directly address this question in the main paper, we include additional results in [Appendix I](appendix_I.html#log-normal-distributions-with-equal-variances) showing that as long as distribution shapes differ across levels, ART tends to confound effects. Naturally, as the shapes become more similar, the problem diminishes. In conclusion, heteroscedasticity may exacerbate the problem, but it is not its sole cause.

>>>>

We will show that ART confounds effects under unequal variances regardless of whether the underlying distribution is skewed. Thus, the problem also arises under normal distributions.
